The Impact of Alternation

### 2017-05-03 draft.tex
ok

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is a combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$. We denote by
$\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$ and write
$\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, we denote by $\phi[t/x]$ the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$. A formula is in positive normal form (PNF) if negation
occurs only within literals and in disjunctive normal form (DNF) if it
is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
\bigwedge_{j=1}^{M_i} \ell_{ij}$. The size of a formula $\phi$,
denoted $\len{\phi}$ is the number of symbols occuring in it.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\Bool^\I$ containing the boolean
  values, and
%
\item each function symbol $f$ into a total function $f^\I :
  \left(\Data^\I\right)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$. Given
formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
\eqiff^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
We omit mentioning the interpretation $\I$ when it is clear from the
context.

@@

In the rest of this section we fix an interpretation $\I$ and a finite
alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
set of \emph{data symbols}. A \emph{data word} (or, simply, a word) is
a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
$\Sigma[\vec{x}]^*$ the set of data words over $\vec{x}$. This
definition extends the classical notion of words from a finite
alphabet with a possibly infinite alphabet $\Sigma[\vec{x}]$. Clearly,
when $\Data^\I$ is sufficiently large or infinite, we can map the
elements of $\Sigma$ into designated elements of $\Data^\I$ and use a
special variable to encode the input events. However, keeping $\Sigma$
explicit in the following simplifies several technical points below,
without cluttering the presentation.

@@

In this section we tackle the following problems: \begin{compactitem}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, does there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and $L(\overline{\A_1})
  = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactitem}

@@

\begin{proposition}\label{prop:overline}
  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
  mapping each $q \in Q$ a value $\nu(q) \in \Bool^\I$ and each $x \in
  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
  $\I,\nu \models \phi \iff \I,\nu' \not\models \overline{\phi}$. 
\end{proposition}
\proof{Immediate, by induction on the structure of $\phi$. \qed}

@@

\section{Predicate Abstraction for Emptiness}

@@

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
finite input alphabet is $\Sigma$, for the rest of this section. Given
a finite sequence $u = a_1\ldots a_n \in \Sigma^*$ of input events and
$\phi \in \Form^+(Q,\vec{x})$, we define $\Post(\phi,u) \equiv \exists
\vec{x}_0 \ldots \exists \vec{x}_{n-1} ~.~
\Delta(\phi,u)[\vec{x}/\vec{x}_n]$.

>>>>

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is a combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
\rightarrow \psi$ for $\neg\phi \vee \psi$.

We denote by $\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$
and write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, we denote by $\phi[t/x]$ the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$. A formula is in positive normal form (PNF) if negation
occurs only within literals and in disjunctive normal form (DNF) if it
is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
\bigwedge_{j=1}^{M_i} \ell_{ij}$. The size of a formula $\phi$,
denoted $\len{\phi}$ is the number of symbols occuring in it.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  \left(\Data^\I\right)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$. Given
formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
\eqiff^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
We omit mentioning the interpretation $\I$ when it is clear from the
context.

@@

In the rest of this section we fix an interpretation $\I$ and a finite
alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
set of \emph{data symbols}. A \emph{data word} (or, simply, a word) is
a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
words over $\vec{x}$. This definition extends the classical notion of
words from a finite alphabet with a possibly infinite alphabet
$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
infinite, we can map the elements of $\Sigma$ into designated elements
of $\Data^\I$ and use a special variable to encode the input
events. However, keeping $\Sigma$ explicit in the following simplifies
several technical points below, without cluttering the presentation.

@@

In this section we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, does there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and $L(\overline{\A_1})
  = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

@@

\begin{proposition}\label{prop:overline}
  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
  mapping each $q \in Q$ a value $\nu(q) \in \booli$ and each $x \in
  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
  $\I,\nu \models \phi \iff \I,\nu' \not\models \overline{\phi}$. 
\end{proposition}
\proof{Immediate, by induction on the structure of $\phi$. \qed}

@@

\section{Antichains for Emptiness}

@@

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define $\Post{\A}(\phi,a) \equiv \exists
\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in
\Form^+(Q,\vec{x})$. We generalize $\Post{\A}$ to finite sequences of
input events: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
exist a sequence $u \in \Sigma^*$ such that the formula $\Accept{\A}(u)
\equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in F} \top \rightarrow
q \wedge \bigwedge_{q \in Q \setminus F} q \rightarrow \bot$ is
satisfiable ? A na\"ive semi-algorithm then enumerates all finite
sequences and checks for the satisfiability of $\Accept{\A}(u)$ for
each $u \in \Sigma^*$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
antichain\footnote{Given a partial order $(D,\preceq)$ an antichain is
  a set $A \subseteq D$ such that $a \not\preceq b$ for any $a,b \in
  A$.} w.r.t. the entailment partial order. This is because, for any
$u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models \Post{\A}(\iota,v)$
and $\Accept{\A}(uw)$ is satisfiable, then $\Accept{\A}(vw)$ is
satisfiable as well, and there is no need to store
$\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be proved
using $v$ alone. However, in practice, the semi-algorithm using
antichains diverges in a large number of cases, because, in many
interpretations, there exist infinite antichains\footnote{For
  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq
  0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u) \mid u \in U}$, where
$U$ is an infinite subset of $\Sigma^*$.

A \emph{safety invariant} for $\A$ is a function $I : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, the following
hold: \begin{compactitem}
\item if $\beta \models \iota$ then $\nu \in I(\beta)$ iff
  $\nu(x)=\mathbf{0}$ for all $x \in \vec{x}$,
%
\item for each $u \in \Sigma^*$, if $\beta \cup \nu \models
  \Post{\A}(\iota,u)$ then $\nu \in I(\beta)$, and
%
\item 
\end{compactitem}
